---
title: "EDA Final project"
author: "mroe1"
date: "28/05/2020"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: united
    pandoc_args: 
      ["--number-sections",
      "--number-offset=1"]
---

```{r, include = FALSE}
library(data.table)
library(ggplot2)
library(corrplot)
library(sp)
library(leaflet)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory data analysis 
In this project, I inversitgated over 150 river stations in europe, and issuses relating to changes in seasonal discharge, particularly if changes in mean runoff have taken place over the last 40 years


## analysis of stations

### station location
stations are located in europe, with particularly density of location in mainland western europe and scandinavia, there are some stations sparely distributed in Russia and eastern europe, and the British Isles and Iceland.
```{r, echo = FALSE}
info_for_project <- readRDS(file = './data/Data for final project-20200428/runoff_eu_info.rds')
df <- data.frame(longitude = info_for_project$Lon, 
                 latitude = info_for_project$Lat)

coordinates(df) <- ~longitude+latitude
leaflet(df) %>% addMarkers(lng = info_for_project$Lon, lat = info_for_project$Lat, popup = paste("Region", info_for_project$Continent, "<br>",
                           "river:", info_for_project$River, "<br>",
                           "country:", info_for_project$Country, "<br>",
                           "altitude", info_for_project$Alt)) %>% addTiles()
```

### number of stations per country/river and number of stations per river
```{r, echo=FALSE}
knitr::kable(info_for_project[, .N, by = Country], caption = "number of stations per country")
knitr::kable(info_for_project[, .N, by = .(River,Country)][, .N, by = Country], caption = "number of rivers per country")
knitr::kable(info_for_project[, .N, by = River], caption = "number of stations per river")
```

### Distribution in space of stations
There is a large cluster of stations around 45-50 degrees in latitude and 5 to 15 degrees in longitude, these stations would correspond to rivers mainly flowing out of alpine regions in Switzerland and Germany.
```{r, echo=FALSE, warning=FALSE}
lon_lat_alt <- ggplot(data = info_for_project, aes(x = Lat, y = Lon, size = Alt)) +
  geom_point() +
  ggtitle("Plot of geographical location and altitude of stations")
lon_lat_alt
lon_lat_alt + coord_cartesian(xlim = c(45, 55), ylim = c(0, 20))+
  ggtitle("Plot of geographical location and altitude of stations--high density of stations(alps)")
theme_set(theme_bw())
```

### distribution by number of years of recorded data
while this is from information given, the actual raw dataset may differ.
```{r, echo=FALSE, warning = FALSE, message = FALSE}
ggplot(info_for_project, aes(x = 1:208, y = N.Years)) + 
  geom_bar(stat = "identity", width = .5, fill = "tomato3") + 
  labs(title = "Number of years of observation", 
       subtitle = "distribution of recorder data") +
  theme(axis.text.x = element_text(angle = 65, vjust = 0.6))
```

## annual and seasonal runoff
### descriptive statistics and ratio of mean high/ mean low values
```{r, echo=FALSE}
daily_runoff <- readRDS(file = './data/Data for final project-20200428/runoff_eu_day.rds')
annual_runoff <- readRDS(file = './data/Data for final project-20200428/runoff_eu_year.rds')
info_for_project <- readRDS(file = './data/Data for final project-20200428/runoff_eu_info.rds')
daily_runoff <- readRDS(file = "./data/daily_runoff.rds")

#daily_runoff$month <- format(as.Date(daily_runoff$date), "%m")
#daily_runoff$year <- format(as.Date(daily_runoff$date), "%Y")
#savedaily_runoff <- saveRDS(daily_runoff, file = "./data/daily_runoff.rds")

daily_runoff <- daily_runoff[value >= 0]
daily_runnoff_descriptive_statistics <- daily_runoff[, .(mean(value), max(value), min(value), sd(value), median(value)), by = .(id,year)]
colnames(daily_runnoff_descriptive_statistics) <-  c("id", "year", "mean", "max", "min", "sd", "median")
daily_runnoff_descriptive_statistics$COF <- daily_runnoff_descriptive_statistics$sd/daily_runnoff_descriptive_statistics$mean
daily_runnoff_descriptive_statistics$skew <- 3 * (daily_runnoff_descriptive_statistics$mean - daily_runnoff_descriptive_statistics$median)/daily_runnoff_descriptive_statistics$sd
daily_runnoff_descriptive_statistics$meanmaxratio <- daily_runnoff_descriptive_statistics$mean/daily_runnoff_descriptive_statistics$max
daily_runnoff_descriptive_statistics$meanminratio <- daily_runnoff_descriptive_statistics$mean/daily_runnoff_descriptive_statistics$min
daily_runnoff_descriptive_statistics$range <- daily_runnoff_descriptive_statistics$max - daily_runnoff_descriptive_statistics$min
characersitic_of_station <- daily_runnoff_descriptive_statistics[, .(max(range), min(range)), by = id]
colnames(characersitic_of_station) <- c("ID", "max_range", "min_range")
characersitic_of_station$ID <- as.numeric(characersitic_of_station$ID)
characersitic_of_station_2 <- merge(characersitic_of_station,info_for_project)
characersitic_of_station_2 <- characersitic_of_station_2[, c(1, 2, 3, 10) ]

```

### catergorisation of the data
After calculating the minimum and maximum run off for each year for each station, I plotted the maximum value and minimum value and made categories for the data. I also categorised by altitude, though some records were missing.
```{r, echo=FALSE, warning=FALSE}
ggplot(characersitic_of_station_2, aes(x=1:202, y = max_range)) +
  geom_point()+
  ggtitle("Plot of max discharge range of stations")
# data could be splti to 0:2500 = low, 2500:5000 = midian, >5000 = high
ggplot(characersitic_of_station_2, aes(x=1:202, y = min_range)) +
  geom_point() + 
  ggtitle("Plot of min discharge range of stations")
# data could be split to 0:500 = low,   500:1000 = median, >1000 high
ggplot(characersitic_of_station_2, aes(x=1:202, y = Alt)) +
  geom_point() +
  ggtitle("Plot of altitude of stations")
# data could be split 0:200 = low, 200:600 = mediam, >600 = high
characersitic_of_station_2[max_range < 2500, max_range_class := factor('low')]
characersitic_of_station_2[max_range > 2500 & max_range < 5000, max_range_class := factor('medium')]
characersitic_of_station_2[max_range > 5000, max_range_class := factor('high')]
characersitic_of_station_2[min_range < 500, min_range_class := factor('low')]
characersitic_of_station_2[min_range > 500 & min_range < 1000, min_range_class := factor('medium')]
characersitic_of_station_2[min_range > 1000, min_range_class := factor('high')]
characersitic_of_station_2[Alt < 200, alt_range_class := factor('low')]
characersitic_of_station_2[Alt > 200 & Alt < 600, alt_range_class := factor('medium')]
characersitic_of_station_2[Alt > 600, alt_range_class := factor('high')]
daily_runnoff_descriptive_statistics[year >= 1980, year_class := factor('Post 1980')]
daily_runnoff_descriptive_statistics[year < 1980, year_class := factor('Pre 1980')]
```

### percantage change in ratio
calculating the mean of the ratios of max/mean and min/mean, for the period before 1980 and after 1980, then calculating the difference between values shows changes in runoff post 1980 relative to pre 1980
```{r, echo=FALSE}
change_in_ratio <- daily_runnoff_descriptive_statistics[, .(mean(meanmaxratio), mean(meanminratio)), by = .(id, year_class)]
change_in_ratio <- change_in_ratio[-1,]
change_in_ratio <- change_in_ratio[-237,]
meanmax_difference <- c()
for (i in 1:length(rownames(change_in_ratio))) {
  meanmax_difference[i] <- change_in_ratio$V1[(2 * i)] - change_in_ratio$V1[((2 * i)-1)]
}
meanmax_difference <- meanmax_difference[1:200]
meanmin_difference <- c()
for (i in 1:length(rownames(change_in_ratio))) {
  meanmin_difference[i] <- change_in_ratio$V1[(2 * i)] - change_in_ratio$V1[((2 * i)-1)]
}
meanmin_difference <- meanmin_difference[1:200]
results <- data.table(meanmax_difference, meanmin_difference, unique(change_in_ratio[,1]))
colnames(results) <- c("meanmax_difference", "meanmin_difference", "ID")
characersitic_of_station_2 <- characersitic_of_station_2[-c(1,120),]
results$ID <- as.numeric(results$ID)
results_2 <- merge(characersitic_of_station_2,results)
results_2 <- results_2[,c(1,5,6,7,8,9)]
knitr::kable(results_2, caption = "class of each station")
```

### aggregation of daily runnoff by season
calculating changes in runoff on a seasonal basis, looking at summer and winter runoff
```{r,echo=FALSE}
daily_runoff$month <- as.numeric(daily_runoff$month)
daily_runoff[(month < 4), season := factor('winter')]
daily_runoff[(month > 6) & (month < 10), season := factor('summer')]
summer_winter_runnoff <- daily_runoff[,mean(value), by = .(id,season)]
a <- which(summer_winter_runnoff$season == "winter")
b <- which(summer_winter_runnoff$season == "summer")
daily_runoff$year <- as.numeric(daily_runoff$year)
daily_runoff[year >= 1980, year_class := factor('Post 1980')]
daily_runoff[year < 1980, year_class := factor('Pre 1980')]
summer_winter <- daily_runoff[,mean(value), by = .(id,season,year_class)]
d <- which(summer_winter$season == "winter")
e <- which(summer_winter$season == "summer")
summer_winter <- summer_winter[c(d, e),]
summer_winter <- summer_winter[-1,]
summer_winter <- summer_winter[-237,]
summer_winter <- summer_winter[-401,]
summer_winter <- summer_winter[-637,]
knitr::kable(summer_winter, caption = "investigating changes in seasonal runnoff pre and post 1980")
```

### percentage change
plotting reulsts, in winter, runnoff has increased over the period 1980-present, this relationship is unclear in the summer period with different rivers showing falls and rises, further investigations would be needed to indentify why this result exists
```{r, echo=FALSE, fig.height=15, warning=FALSE}
percentage_change <- c()
for (i in 1:length(rownames(summer_winter))) {
  percentage_change[i] <- 100*(summer_winter$V1[(2 * i)] - summer_winter$V1[((2 * i)-1)])/(summer_winter$V1[((2 * i)-1)])
}
winter <- percentage_change[1:200]
summer <- percentage_change[201:400]
results_3 <- data.table(unique(summer_winter[,1]), winter, summer)
results_3$ID_type_winter <- ifelse(winter < 0, "below", "above")
ggplot(results_3, aes(x=id, y=winter, label=winter)) + 
  geom_bar(stat='identity', aes(fill=ID_type_winter), width=.5)  +
                    scale_fill_manual(name="percentage change", 
                                      labels = c("Positive", "Negitive"), 
                                      values = c("above"="#00ba38", "below"="#f8766d")) + 
                      labs(subtitle="percentage change of mean runnoff in Winter at stations before and after 1980", 
                           title= "Diverging Bars") + 
                      coord_flip()
results_3$ID_type_summer <- ifelse(summer < 0, "below", "above")
ggplot(results_3, aes(x=id, y=summer, label=summer)) + 
  geom_bar(stat='identity', aes(fill=ID_type_summer), width=.5)  +
  scale_fill_manual(name="percentage change", 
                    labels = c("Positive", "Negitive"), 
                    values = c("above"="#00ba38", "below"="#f8766d")) + 
  labs(subtitle="percentage change of mean runnoff at stations in summer before and after 1980", 
       title= "Diverging Bars") + 
  coord_flip()
```

## Five chsoen rivers
### Five rivers
Danube river-bratislava,
while there are no dams between Vienna and Bratislava, the danube has many dams in its upper channel and just past the station there is a dam, there has been lots of conflict between Slovakia and hungary on wheter further interventions should take place

rhine – upper parts of the rhine have many dams
sitter – 11 hydroelectric plants along the river 
LUETSCHINE - a small river heading to a lake, in high altitude influenced heavily by snowmelt
glama – many power stations have been built along the river, in Sweden.

I chose stations based on large mean values, large standard deviations, large altitude, high latitude, and high max values, were duplicate rivers were chosen, another station which a smaller value was chosen
```{r, echo=FALSE}
number_of_entries <- daily_runoff[, .N, by = id]
number_of_entries <- number_of_entries[N > 36500]
daily_runoff <- daily_runoff[value >= 0]
number_of_entries <- daily_runoff[, .N, by = id]
number_of_entries <- number_of_entries[N > 36500]
analysis_of_means <- daily_runnoff_descriptive_statistics[, mean(mean), by = id]
test <- daily_runnoff_descriptive_statistics[, mean(sd), by = id][order(-V1)][1:12,1]
five_rivers <- data.table()
high_means <- analysis_of_means[order(-V1)][c(1:20),1]
five_rivers$id <- high_means[20]
sd <- daily_runnoff_descriptive_statistics[, mean(sd), by = id]
five_rivers$id2 <- sd[order((-V1))][1:10, 1][9]
five_rivers$id3 <- daily_runnoff_descriptive_statistics[, mean(max), by = id][order(-V1)][5,1]
five_rivers$id4 <- info_for_project[,.(ID,Lat)][order(-Lat)][201,1]
five_rivers$id5 <- info_for_project[,.(ID,Alt)][order(-Alt)][5,1]
chosen_stations <- info_for_project[ID %in% five_rivers]
knitr::kable(chosen_stations, caption = "chosen stations")
```

### Seasonal runnoff at monthly scale
3 of the rivers, on the rhine, the sitter and the danube, show flat profiles over months (balanced between greater precipitaiton in the winter and more snowmelt during summer), river glama is probably heavly influeced by snow melt due to large rises in runnoff in spring and summer months, as is the Luetshine river which, is a short river at high altitude.
```{r, echo=FALSE,  warning = FALSE, message = FALSE}
position_in_annual_runoff <- daily_runoff[id == chosen_stations$ID[1] | id == chosen_stations$ID[2] | id == chosen_stations$ID[3] | id == chosen_stations$ID[4] | id == chosen_stations$ID[5]]
monthly_average_discharge <- position_in_annual_runoff[,mean(value) ,by = .(id,month,year)]

ggplot(monthly_average_discharge, aes(x = factor(month), y = V1, group = month)) +
  geom_boxplot() +
  facet_wrap(~id, scales = 'free') +
  xlab(label = "Month") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle("montlhy runoff of chosen stations") +
  theme_bw() 
  
```

### correlation matrix
glama river is highly uncorrelated with the other rivers.
```{r, echo=FALSE, message=FALSE}
yearly_annual_discharge_5_rivers <- position_in_annual_runoff[, sum(value), by = .(id,year)]
runnoff_year_mat <- dcast(yearly_annual_discharge_5_rivers, year~id)
runnoff_year_mat <- runnoff_year_mat[97:193,]
runoff_year_cor <- cor(runnoff_year_mat[, -1], use = "pairwise.complete.obs")
runoff_year_cor <- round(runoff_year_cor,2)

col1 <- colorRampPalette(c("#7F0000", "red", "#FF7F00", "yellow", "white",
                           "cyan", "#007FFF", "blue", "#00007F"))
col2 <- colorRampPalette(c("#67001F", "#B2182B", "#D6604D", "#F4A582",
                           "#FDDBC7", "#FFFFFF", "#D1E5F0", "#92C5DE",
                           "#4393C3", "#2166AC", "#053061"))
col3 <- colorRampPalette(c("red", "white", "blue")) 
col4 <- colorRampPalette(c("#7F0000", "red", "#FF7F00", "yellow", "#7FFF7F",
                           "cyan", "#007FFF", "blue", "#00007F"))
whiteblack <- c("white", "black")
corrplot(runoff_year_cor, order = "hclust", addrect = 2, col = col4(10), method = "number", title = "similarity between stations")
```

### regreesion during summer and winter season of chosen stations
winter runoff is increasing as a function of time- espcially at river glama. In summer runnoff, runnoff is faling except for station luetschine
```{r, echo=FALSE}
colset_4 <-  c("#D35C37", "#BF9A77", "#D6C6B9", "#97B8C2")
runoff_5_rivers_2 <- position_in_annual_runoff
#runoff_5_rivers_2
runoff_5_rivers_2$month <- as.numeric(factor((runoff_5_rivers_2$month)))
runoff_5_rivers_2[month == 1 | month == 2 | month == 3, season := factor('winter')]
runoff_5_rivers_2[month == 7 | month == 8 | month == 9, season := factor('summer')]


summer_runnof <- runoff_5_rivers_2[season == "summer" ]
summer_runnof <- summer_runnof[,sum(value), by = .(year,id)]
winter_runnof <- runoff_5_rivers_2[season == "winter" ]
winter_runnof <- winter_runnof[,sum(value), by = .(year,id)]
summer_runnof$year <- as.numeric(factor((summer_runnof$year)))
winter_runnof$year <- as.numeric(factor((winter_runnof$year)))


ggplot(summer_runnof, aes(x = year, y = V1)) +
  geom_line(col = colset_4[3], aes(group = 1))+
  geom_point(col = colset_4[3])+
  facet_wrap(~id, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 3, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle("summer runoff regression plots for chosen stations") +
  theme_bw()
ggplot(winter_runnof, aes(x = year, y = V1)) +
  geom_line(col = colset_4[3], aes(group = 1))+
  geom_point(col = colset_4[3])+
  facet_wrap(~id, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 3, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle("winter runoff regression plots for chosen stations") +
  theme_bw()
```

### boxplot analysis
these plots also indicate a clear trend of increasing winter runnoff, while inconclusive evidence on summer runnoff
```{r, fig.height=20,fig.width=20, echo=FALSE}
daily_runoff <- readRDS(file = "./data/daily_runoff.rds")
all_data_100plusyears <- daily_runoff[id %in% number_of_entries$id]
all_data_100plusyears$month <- as.numeric(factor(all_data_100plusyears$month))
all_data_100plusyears[month == 01 | month == 02 | month == 03, season := factor('winter')]
all_data_100plusyears[month == 07 | month == 08 | month == 09, season := factor('summer')]
all_data_100plusyears[month == 04 | month == 05 | month == 06, season := factor('spring')]
all_data_100plusyears[month == 10 | month == 11 | month == 12, season := factor('autumn')]
all_data_100plusyears[, value_norm := scale(value), by = .(id, year, season)]
yearly_annual_data <- all_data_100plusyears[, sum(value), by = .(id,season,year)]
yearly_annual_data[, value_norm := scale(V1), by = .(id,season)]
summer <- yearly_annual_data[season == "summer"]
winter <- yearly_annual_data[season == "winter"]

winter2 <- winter[year > 1950]
summer2 <- summer[year > 1950]
year_thres <- 1980
to_plot <- rbind(cbind(winter2, season = factor('winter')), 
                 cbind(summer2, season = factor('summer'))) 

to_plot[year < year_thres, period := factor('1950-1980')]
to_plot[year >= year_thres, period := factor('1981-2016')]
to_plot[year < year_thres, period := factor('1950-1980')]
to_plot[year >= year_thres, period := factor('1981-2016')]

to_plot <- to_plot[year >= 1950]
to_plot <- to_plot[id %in% unique(to_plot$id)]
colnames(to_plot) <- make.unique(names(to_plot))

ggplot(to_plot, aes(season.1, V1, fill = period)) +
  geom_boxplot() +
  facet_wrap(~id, scales = 'free_y') +
  xlab(label = "Season") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle("box plots for all stations") +
  theme_bw()

to_plot2 <- to_plot[id %in% chosen_stations$ID]
ggplot(to_plot2, aes(season.1, V1, fill = period)) +
  geom_boxplot() +
  facet_wrap(~id, scales = 'free_y') +
  xlab(label = "Season") +
  ylab(label = "Runoff (m3/s)") +
  ggtitle("box plots for chosen stations") +
  theme_bw()
```

