---
title: "EDA Final project"
author: "mroe1"
date: "28/05/2020"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: united
    pandoc_args: 
      ["--number-sections",
      "--number-offset=1"]
---

```{r, include = FALSE}
library(data.table)
library(ggplot2)
library(corrplot)
library(sp)
library(leaflet)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory data analysis

In my exploratory data analysis of the river Rhine, I produced tables and graphs on various data sets of discharge at different stations along the course of the river Rhine.
The anaysis hypothesis, suggests falling summer discharge and increasesing winter discharge, due to climate change and it knock on effects on snow depths in mountains regions and incresed evapotranspiration during summer months.

## analysis of stations
brocken down by various months and various stations along the river, it is clear were data is missing due to -999 value

```{r, echo = FALSE}
info_for_project <- readRDS(file = './data/Data for final project-20200428/runoff_eu_info.rds')
df <- data.frame(longitude = info_for_project$Lon, 
                 latitude = info_for_project$Lat)

coordinates(df) <- ~longitude+latitude
leaflet(df) %>% addMarkers(lng = info_for_project$Lon, lat = info_for_project$Lat, popup = paste("Region", info_for_project$Continent, "<br>",
                           "river:", info_for_project$River, "<br>",
                           "country:", info_for_project$Country, "<br>",
                           "altitude", info_for_project$Alt)) %>% addTiles()
#stations are located in europe, particularly dense in mainland western europe and scandinavia, there are some stations sparely distributed in Russia and eastern europe, and the British Isles and Iceland.
lon_lat_alt <- ggplot(data = info_for_project, aes(x = Lat, y = Lon, size = Alt)) +
  geom_point()
lon_lat_alt
lon_lat_alt + coord_cartesian(xlim = c(45, 55), ylim = c(0, 20))
ggplot(data = info_for_project, aes(x = 1:208, y = N.Years)) +
   geom_point()

theme_set(theme_bw())

# Draw plot
ggplot(info_for_project, aes(x = 1:208, y = N.Years)) + 
  geom_bar(stat="identity", width=.5, fill="tomato3") + 
  labs(title="Number of years of observation", 
       subtitle="distribution of recorder data", 
       caption="source: https://www.bafg.de/GRDC/EN/01_GRDC/grdc_node.html") + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6))
# with a few exceptions most stations have around 100 years of observations
```

## annual and seasonal runoff
This table shows a plots of summary statitics of all stations over the period of all data collected, it shows the spread of the data. This could be important in being able to catagorise stations into specific subsets, such as differences in mean values, or the differences in the range, hence patterens could be deduced. Data with incorrect values from previous table has been removed.

```{r ,echo = FALSE}
daily_runoff <- readRDS(file = './data/Data for final project-20200428/runoff_eu_day.rds')
annual_runoff <- readRDS(file = './data/Data for final project-20200428/runoff_eu_year.rds')
info_for_project <- readRDS(file = './data/Data for final project-20200428/runoff_eu_info.rds')
daily_runoff <- readRDS(file = "./data/daily_runoff.rds")

#daily_runoff$month <- format(as.Date(daily_runoff$date), "%m")
#daily_runoff$year <- format(as.Date(daily_runoff$date), "%Y")
#savedaily_runoff <- saveRDS(daily_runoff, file = "./data/daily_runoff.rds")

daily_runoff <- daily_runoff[value >= 0]
daily_runnoff_descriptive_statistics <- daily_runoff[, .(mean(value), max(value), min(value), sd(value), median(value)), by = .(id,year)]
colnames(daily_runnoff_descriptive_statistics) <-  c("id","year","mean","max","min","sd","median")
daily_runnoff_descriptive_statistics
daily_runnoff_descriptive_statistics$COF <- daily_runnoff_descriptive_statistics$sd/daily_runnoff_descriptive_statistics$mean
daily_runnoff_descriptive_statistics$skew <- 3*(daily_runnoff_descriptive_statistics$mean - daily_runnoff_descriptive_statistics$median)/daily_runnoff_descriptive_statistics$sd
daily_runnoff_descriptive_statistics$meanmaxratio <- daily_runnoff_descriptive_statistics$mean/daily_runnoff_descriptive_statistics$max
daily_runnoff_descriptive_statistics$meanminratio <- daily_runnoff_descriptive_statistics$mean/daily_runnoff_descriptive_statistics$min
daily_runnoff_descriptive_statistics
daily_runnoff_descriptive_statistics$range <- daily_runnoff_descriptive_statistics$max - daily_runnoff_descriptive_statistics$min
characersitic_of_station <- daily_runnoff_descriptive_statistics[, .(max(range), min(range)), by = id]
colnames(characersitic_of_station) <- c("ID","max_range", "min_range")
characersitic_of_station$ID <- as.numeric(characersitic_of_station$ID)
characersitic_of_station_2 <- merge(characersitic_of_station,info_for_project)
characersitic_of_station_2 <- characersitic_of_station_2[,c(1,2,3,10)]
ggplot(characersitic_of_station_2, aes(x=1:202, y = max_range)) +
  geom_point()
# data could be splti to 0:2500 = low, 2500:5000 = midian, >5000 = high
ggplot(characersitic_of_station_2, aes(x=1:202, y = min_range)) +
  geom_point()
# data could be split to 0:500 = low,   500:1000 = median, >1000 high
ggplot(characersitic_of_station_2, aes(x=1:202, y = Alt)) +
  geom_point()
# data could be split 0:200 = low, 200:600 = mediam, >600 = high
characersitic_of_station_2[max_range < 2500, max_range_class := factor('low')]
characersitic_of_station_2[max_range > 2500 & max_range < 5000, max_range_class := factor('medium')]
characersitic_of_station_2[max_range > 5000, max_range_class := factor('high')]
characersitic_of_station_2[min_range < 500, min_range_class := factor('low')]
characersitic_of_station_2[min_range > 500 & min_range < 1000, min_range_class := factor('medium')]
characersitic_of_station_2[min_range > 1000, min_range_class := factor('high')]
characersitic_of_station_2[Alt < 200, alt_range_class := factor('low')]
characersitic_of_station_2[Alt > 200 & Alt < 600, alt_range_class := factor('medium')]
characersitic_of_station_2[Alt > 600, alt_range_class := factor('high')]
daily_runnoff_descriptive_statistics[year >= 1980, year_class := factor('Post 1980')]
daily_runnoff_descriptive_statistics[year < 1980, year_class := factor('Pre 1980')]
change_in_ratio <- daily_runnoff_descriptive_statistics[, .(mean(meanmaxratio), mean(meanminratio)), by = .(id, year_class)]
change_in_ratio <- change_in_ratio[-1,]
change_in_ratio <- change_in_ratio[-237,]
meanmax_difference <- c()
for (i in 1:length(rownames(change_in_ratio))) {
  meanmax_difference[i] <- change_in_ratio$V1[(2 * i)] - change_in_ratio$V1[((2 * i)-1)]
}
meanmax_difference <- meanmax_difference[1:200]
meanmin_difference <- c()
for (i in 1:length(rownames(change_in_ratio))) {
  meanmin_difference[i] <- change_in_ratio$V1[(2 * i)] - change_in_ratio$V1[((2 * i)-1)]
}
meanmin_difference <- meanmin_difference[1:200]
results <- data.table(meanmax_difference, meanmin_difference, unique(change_in_ratio[,1]))
colnames(results) <- c("meanmax_difference", "meanmin_difference", "ID")
characersitic_of_station_2 <- characersitic_of_station_2[-c(1,120),]
results$ID <- as.numeric(results$ID)
results_2 <- merge(characersitic_of_station_2,results)
results_2 <- results_2[,c(1,5,6,7,8,9)]
daily_runoff$month <- as.numeric(daily_runoff$month)
daily_runoff[(month < 4), season := factor('winter')]
daily_runoff[(month > 6) & (month < 10), season := factor('summer')]
summer_winter_runnoff <- daily_runoff[,mean(value), by = .(id,season)]
a <- which(summer_winter_runnoff$season == "winter")
b <- which(summer_winter_runnoff$season == "summer")
daily_runoff$year <- as.numeric(daily_runoff$year)
daily_runoff[year >= 1980, year_class := factor('Post 1980')]
daily_runoff[year < 1980, year_class := factor('Pre 1980')]
summer_winter <- daily_runoff[,mean(value), by = .(id,season,year_class)]
d <- which(summer_winter$season == "winter")
e <- which(summer_winter$season == "summer")
summer_winter <- summer_winter[c(d, e),]
summer_winter <- summer_winter[-1,]
summer_winter <- summer_winter[-237,]
summer_winter <- summer_winter[-401,]
summer_winter <- summer_winter[-637,]
percentage_change <- c()
for (i in 1:length(rownames(summer_winter))) {
  percentage_change[i] <- 100*(summer_winter$V1[(2 * i)] - summer_winter$V1[((2 * i)-1)])/(summer_winter$V1[((2 * i)-1)])
}
winter <- percentage_change[1:200]
summer <- percentage_change[201:400]
results_3 <- data.table(unique(summer_winter[,1]), winter, summer)
results_3$ID_type_winter <- ifelse(winter < 0, "below", "above")
ggplot(results_3, aes(x=id, y=winter, label=winter)) + 
  geom_bar(stat='identity', aes(fill=ID_type_winter), width=.5)  +
                    scale_fill_manual(name="percentage change", 
                                      labels = c("Positive", "Negitive"), 
                                      values = c("above"="#00ba38", "below"="#f8766d")) + 
                      labs(subtitle="percentage change of mean runnoff at stations before and after 1980", 
                           title= "Diverging Bars") + 
                      coord_flip()
                    
  coord_flip()
results_3$ID_type_summer <- ifelse(summer < 0, "below", "above")
ggplot(results_3, aes(x=id, y=summer, label=summer)) + 
  geom_bar(stat='identity', aes(fill=ID_type_summer), width=.5)  +
  scale_fill_manual(name="percentage change", 
                    labels = c("Positive", "Negitive"), 
                    values = c("above"="#00ba38", "below"="#f8766d")) + 
  labs(subtitle="percentage change of mean runnoff at stations before and after 1980", 
       title= "Diverging Bars") + 
  coord_flip()
```

## five rivers selected
To show and identify points in time when maximum points and minimum points were reached in the stations run off 
data, I created interactive plots of each station, it is also easy to see if there is correlation between station or not. It can also be seen were data is missing.

```{r, echo = FALSE}
colset_4 <-  c("#D35C37", "#BF9A77", "#D6C6B9", "#97B8C2")

daily_runoff <- daily_runoff[value >= 0]

number_of_entries <- daily_runoff[, .N, by = id]
number_of_entries <- number_of_entries[N > 36500]



daily_runoff$month <- as.numeric(factor((daily_runoff$month)))
daily_runoff[month == 1 | month == 2 | month == 3, season := factor('winter')]
daily_runoff[month == 7 | month == 8 | month == 9, season := factor('summer')]
all_summer_runoff <- daily_runoff[season == "summer" ]  
all_summer_runoff <- all_summer_runoff[,sum(value), by = .(year,id)]
ggplot(all_summer_runoff, aes(x = year, y = V1)) +
  geom_line(col = colset_4[3], aes(group = 1))+
  geom_point(col = colset_4[3])+
  facet_wrap(~id, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 3, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()

#high mean value, #high sd, #high max, #high long, #high altitude
# take 20 random samples(so assess similarity to other data)
#==============================
daily_runoff <- daily_runoff[value >= 0]
number_of_entries <- daily_runoff[, .N, by = id]
number_of_entries <- number_of_entries[N > 36500]
analysis_of_means <- daily_runnoff_descriptive_statistics[, mean(mean), by = id]
test <- daily_runnoff_descriptive_statistics[, mean(sd), by = id][order(-V1)][1:12,1]
analysis_of_means[order(-V1)]
five_rivers <- data.table()
high_means <- analysis_of_means[order(-V1)][c(1:10),1]
five_rivers$id <- high_means[6]
sd <- daily_runnoff_descriptive_statistics[, mean(sd), by = id]

merge(number_of_entries, sd[order((-V1))][1:20, 1])
five_rivers$id2 <- sd[order((-V1))][1:10, 1][9]


five_rivers$id3 <- daily_runnoff_descriptive_statistics[, mean(max), by = id][order(-V1)][5,1]
five_rivers$id4 <- info_for_project[,.(ID,Lat)][order(-Lat)][201,1]

five_rivers$id5 <- info_for_project[,.(ID,Alt)][order(-Alt)][5,1]

daily_runoff[id %in% five_rivers]
chosen_stations <- info_for_project[ID %in% five_rivers]


position_in_annual_runoff <- daily_runoff[id == chosen_stations$ID[1] | id == chosen_stations$ID[2] | id == chosen_stations$ID[3] | id == chosen_stations$ID[4] | id == chosen_stations$ID[5]]


monthly_average_discharge <- position_in_annual_runoff[,mean(value) ,by = .(id,month,year)]
monthly_average_discharge
ggplot(monthly_average_discharge, aes(x = factor(month), y = V1, group = month)) +
  geom_boxplot() +
  facet_wrap(~id, scales = 'free') +
  xlab(label = "Month") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
yearly_annual_discharge_5_rivers <- position_in_annual_runoff[, sum(value), by = .(id,year)]
runnoff_year_mat <- dcast(yearly_annual_discharge_5_rivers, year~id)
runnoff_year_mat <- runnoff_year_mat[97:193,]
runnoff_year_mat
runoff_year_cor <- cor(runnoff_year_mat[, -1], use = "pairwise.complete.obs")
runoff_year_cor <- round(runoff_year_cor,2)

col1 <- colorRampPalette(c("#7F0000", "red", "#FF7F00", "yellow", "white",
                           "cyan", "#007FFF", "blue", "#00007F"))
col2 <- colorRampPalette(c("#67001F", "#B2182B", "#D6604D", "#F4A582",
                           "#FDDBC7", "#FFFFFF", "#D1E5F0", "#92C5DE",
                           "#4393C3", "#2166AC", "#053061"))
col3 <- colorRampPalette(c("red", "white", "blue")) 
col4 <- colorRampPalette(c("#7F0000", "red", "#FF7F00", "yellow", "#7FFF7F",
                           "cyan", "#007FFF", "blue", "#00007F"))
whiteblack <- c("white", "black")
corrplot(runoff_year_cor, order = "hclust", addrect = 2, col = col4(10), method = "number")
#runoff_5_rivers <- runoff_5_rivers[,-6]
#runoff_5_rivers
runoff_5_rivers_2 <- position_in_annual_runoff
#runoff_5_rivers_2
runoff_5_rivers_2$month <- as.numeric(factor((runoff_5_rivers_2$month)))
runoff_5_rivers_2[month == 1 | month == 2 | month == 3, season := factor('winter')]
runoff_5_rivers_2[month == 7 | month == 8 | month == 9, season := factor('summer')]


summer_runnof <- runoff_5_rivers_2[season == "summer" ]
summer_runnof <- summer_runnof[,sum(value), by = .(year,id)]
winter_runnof <- runoff_5_rivers_2[season == "winter" ]
winter_runnof <- winter_runnof[,sum(value), by = .(year,id)]
summer_runnof$year <- as.numeric(factor((summer_runnof$year)))
winter_runnof$year <- as.numeric(factor((winter_runnof$year)))


ggplot(summer_runnof, aes(x = year, y = V1)) +
  geom_line(col = colset_4[3], aes(group = 1))+
  geom_point(col = colset_4[3])+
  facet_wrap(~id, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 3, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
ggplot(winter_runnof, aes(x = year, y = V1)) +
  geom_line(col = colset_4[3], aes(group = 1))+
  geom_point(col = colset_4[3])+
  facet_wrap(~id, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = colset_4[1]) +
  geom_smooth(method = 'loess', formula = y~x, se = 0, col = colset_4[4]) +
  scale_color_manual(values = colset_4[c(1, 2, 3, 4)]) +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
all_data_100plusyears <- daily_runoff[id %in% number_of_entries$id]
all_data_100plusyears$month <- as.numeric(factor(all_data_100plusyears$month))
all_data_100plusyears[month == 01 | month == 02 | month == 03, season := factor('winter')]
all_data_100plusyears[month == 07 | month == 08 | month == 09, season := factor('summer')]
all_data_100plusyears[month == 04 | month == 05 | month == 06, season := factor('spring')]
all_data_100plusyears[month == 10 | month == 11 | month == 12, season := factor('autumn')]
all_data_100plusyears[, value_norm := scale(value), by = .(id, year, season)]
yearly_annual_data <- all_data_100plusyears[, sum(value), by = .(id,season,year)]
yearly_annual_data[, value_norm := scale(V1), by = .(id,season)]
all_data_100plusyears[, mean(value), by = .(id,year,season)]
summer <- yearly_annual_data[season == "summer"]
n_stations <- unique(summer$id)

summer_test <- summer[id %in% n_stations[1:20]]
low_altittude <- (info_for_project[order(-Alt)])
low_alt_stations <- unique(low_altittude[120:165,]$ID)
test <- unique(summer[id %in% low_alt_stations]$id)
summer_test <- summer[id %in% test]
ggplot(summer_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
summer_test[year == "2009"][value_norm > 0.2]$id
low_altittude <- (info_for_project[order(-Alt)])
high_alt_stations <- unique(low_altittude[1:50,]$ID)
test <- unique(summer[id %in% high_alt_stations]$id)
summer_test <- summer[id %in% test]
ggplot(summer_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()

low_lat <- (info_for_project[order(-Lat)])
low_lat_stations <- unique(low_lat[170:208,]$ID)
test <- unique(summer[id %in% low_lat_stations]$id)
summer_test <- summer[id %in% test]
ggplot(summer_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
low_lat <- (info_for_project[order(-Lat)])
high_lat_stations <- unique(low_lat[1:100,]$ID)
test <- unique(summer[id %in% high_lat_stations]$id)
summer_test <- summer[id %in% test]
ggplot(summer_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
winter <- yearly_annual_data[season == "winter"]
final_year <- summer_test[year == 2009]
final_year[value_norm >= 0.2, level := factor('high')]
final_year[value_norm <= -0.2, level := factor('low')]
final_year[value_norm < 0.2 && value_norm > -0.2, level := factor('medium')]
final_year[level == "high"]$id
info_for_project[ID %in% final_year[level == "high"]$id]
info_for_project[ID %in% final_year[level == "medium"]$id]
info_for_project[ID %in% final_year[level == "low"]$id]
n_stations_winter <- unique(winter$id)
winter_test <- winter[id %in% n_stations[1:20]]
low_altittude <- (info_for_project[order(-Alt)])
low_alt_stations <- unique(low_altittude[120:165,]$ID)
test <- unique(winter[id %in% low_alt_stations]$id)
winter_test <- winter[id %in% test]
ggplot(winter_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()

n_stations_winter <- unique(winter$id)
winter_test <- winter[id %in% n_stations[1:20]]
low_altittude <- (info_for_project[order(-Alt)])
high_alt_stations <- unique(low_altittude[1:50,]$ID)
test <- unique(winter[id %in% high_alt_stations]$id)
winter_test <- winter[id %in% test]
ggplot(winter_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
###########
low_long <- (info_for_project[order(-Lon)])
high_long_stations <- unique(low_lat[1:50,]$ID)
test <- unique(summer[id %in% high_long_stations]$id)
summer_test <- summer[id %in% test]
ggplot(summer_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
low_long <- (info_for_project[order(-Lon)])
low_long_stations <- unique(low_lat[160:208,]$ID)
test <- unique(summer[id %in% low_long_stations]$id)
summer_test <- summer[id %in% test]
ggplot(summer_test[year > 1910 | year < 2000], aes(x = year, y = value_norm, col = id, group = id)) +
  geom_smooth(method = 'loess', formula = y~x, se = 0) + 
  ggtitle('Winter runoff') +
  xlab(label = "Year") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()
winter2 <- winter[year > 1950]
summer2 <- summer[year >1950]
year_thres <- 1980
to_plot <- rbind(cbind(winter2, season = factor('winter')), 
                 cbind(summer2, season = factor('summer'))) 

to_plot[year < year_thres, period := factor('1950-1980')]
to_plot[year >= year_thres, period := factor('1981-2016')]
to_plot[year < year_thres, period := factor('1950-1980')]
to_plot[year >= year_thres, period := factor('1981-2016')]

to_plot <- to_plot[year >= 1950]
to_plot <- to_plot[id %in% unique(to_plot$id)]
colnames(to_plot) <- make.unique(names(to_plot))

ggplot(to_plot, aes(season.1, V1, fill = period)) +
  geom_boxplot() +
  facet_wrap(~id, scales = 'free_y') +
  xlab(label = "Season") +
  ylab(label = "Runoff (m3/s)") +
  theme_bw()


```



